---
title: "Week 5 Homework: Tidying Data"
author: "Aaron Grzasko"
date: "October 2, 2016"
output: 
    html_document:
        theme: flatly
        highlight: haddock
---

---  

## Assignment Overview

The primary purpose of this exercise is to gain experience manipulating untidy data.  We will use two powerful R packages--tidyr and dplyr--to wrangle a small albeit messy data set into tidy form.  
  
According to Hadley Wickham--see [*Tidy Data R Vignette*](ftp://cran.r-project.org/pub/R/web/packages/tidyr/vignettes/tidy-data.html)--tidy data conform to three principles:  
  
* Variables are stored in separate columns
* Observations are stored in separate rows
* Each observational unit forms a table  
  
Once we have tidied our data, we will perform basic graphical and statistical analysis.  
  
## Libraries  
```{r load_lib, warning=FALSE, message=FALSE}
# install.packages(c("tidyr","dplyr"))
library(tidyr)
library(dplyr)
```

## Data  
  
In this assignment, we review flight delay data for two airline carriers and five separate cities--refer to the [assignment instructions](https://bbhosted.cuny.edu/bbcswebdav/pid-25842532-dt-content-rid-119781391_1/xid-119781391_1) for more details.  
 
This data was inputted manually in a .csv file and saved to a [Github repository](https://github.com/spitakiss/Data607_HW5/blob/master/flight_info.csv).  
  
Let's load the csv file into the R environment and view its contents:  
```{r loaddata, comment=NA}
myurl <- "https://raw.githubusercontent.com/spitakiss/Data607_HW5/master/flight_info.csv"
flight <- read.csv(myurl, header = TRUE, stringsAsFactors = FALSE)
flight
```
  
The current `flight` data frame is in a messy, wide format and will require multiple transforms to produce an object in tidy form.  

## Tidy Operations  

In transforming the raw `flight` object, we made the following assumptions:  
  
* The flight statuses "on time" and "delayed" are separate variables and should be stored in separate columns.  
* The various city columns in the raw data set belong to a master *city* variable and should be stored in a single column.
* An observation is defined by flight metrics for a specific airline and city combination.  
  
Here is an overview of the tidying operations performed--also refer to the annotations within the subsequent code chunk:  
  
1.  *Rename columns*.  The original csv file did have proper names for the first two columns: airline or flight status.  By default, the `read.csv()` function named these columns "X"" and "X1".  We use the `rename()` function to update with proper labels.  
2.  *Delete whitespace*.  We use the dplyr function `filter()` to eliminate the unnecessary row that separates observations relating to Alaska and AM West airlines.  
3.  *Update elements in existing fields.*  We made adjustments to specific elements the *airline* and *status* fields:  
    + *Airlines*: Some rows had missing values for the appropriate airline carrier.  Fortunately, the correct carrier information for these rows were located in the preceding record.  Values were populated using the `mutate()`, `ifelse()`, and `lag()` functions.  
    + *Status*: In a later step we treat the element "on time" as separate variable stored in  a unique column.  We replace "on time" with "on_time" to follow standard naming conventions for field names.  
4.  *Convert city columns to rows*.  We assume *city* is a variable in this analysis. Using `gather()`, we convert the relevant city columns to rows.  
5.  *Store flight status elements as separate columns*.  This transformation was achieved using the `spread()` function.
6.  *Add new columns with relevant metrics.*  Using `mutate()`, we added a new column that calculate the total number of flights for given city/airport combination.  We also added two additional columns that describe each observation's on time and delayed flight percentage, respectively.   



```{r comment=NA}
flight <-   flight %>%
                #1
                rename(airline = X, status = X.1) %>%  #1
                #2
                filter(status != "") %>%
                # 3
                mutate(
                    airline  = ifelse(airline == "", lag(airline),airline),
                    status = ifelse(status == "on time", "on_time", status)
                    ) %>%
                # 4
                gather("city", "num_flights", Los.Angeles:Seattle) %>%
                # 5
                spread(status, num_flights) %>%
                # 6
                mutate(
                    ttl_flights = delayed + on_time,
                    on_time_per = round(on_time / ttl_flights,4),
                    delay_per = round(delayed / ttl_flights,4)
                )
    
flight
```

## Graphical and Tabular Summaries  
  
**By City**  
Below we summarize the proportion of delayed flights by city for each airline, with output in both tabular and bar plot formats:  

```{r, comment=NA}

flight.redux <- flight %>%
                    select(c(airline,city, delay_per)) %>%
                    spread(city, delay_per) %>%
                    select(Los.Angeles:Seattle) %>%
                    as.matrix()
rownames(flight.redux) <- unique(flight$airline)

flight.redux

barplot(flight.redux, beside = TRUE, main = "Proportion of Flights Delayed", xlab = "City", col=c("royalblue3","palegreen3"))          
legend("topleft", rownames(flight.redux), pch = 15, bty = 'n', col = c("royalblue3","palegreen3"))
```
  
Interestingly enough, Alaska Airlines has a lower proportion of flight delays than AM West for each of the five observed cities.  
  
**In Aggregate**  
Now, let's produce a summary table and bar plot by carrier for all cities combined:   

```{r, comment=NA}
flight_all_city <- flight %>%
                    group_by(airline) %>%
                    summarise(delay_prop = sum(delayed)/sum(ttl_flights)) %>%
                    select(delay_prop) %>%
                    as.matrix()

rownames(flight_all_city) <- unique(flight$airline)

flight_all_city

barplot(flight_all_city, beside = TRUE, main = "Proportion of Flights Delayed", xlab = "All Cities", col=c("royalblue3","palegreen3"), ylim = c(0,0.2))      

legend("topleft", rownames(flight_all_city), pch = 15, bty = 'n', col = c("royalblue3","palegreen3"))

```
  
Surprisingly, the  flight data--when aggregated together--reveal a striking pattern reversal:  AM West's total flight delay percentage is lower than the corresponding percentage for Alaska!  
  
These counter intuitive results are common to the world of statisticians.  In fact, this phenomenon even has a name:  [Simpson's paradox](https://en.wikipedia.org/wiki/Simpson%27s_paradox).  
  
  
## Statistical Analysis
  
**By City**  
As noted previously, the proportion of delayed flights for Alaska airlines is lower than AM West for each of the five cities in the data set.  Let's now use the function `prop.text()` to determine if the calculated difference in proportions is statistically meaningful at the 5% level of significance.  

```{r, stat_test, comment=NA}
for (i in unique(flight$city)){ 
    flight_sub <- filter(flight, city == i)
    delay_AM <- sum(select(filter(flight_sub, airline == 'AM WEST'),delayed))
    delay_AK <- sum(select(filter(flight_sub, airline =='ALASKA'), delayed))
    ttl_AM <- sum(select(filter(flight_sub, airline == 'AM WEST'),ttl_flights))
    ttl_AK <- sum(select(filter(flight_sub, airline == 'ALASKA'),ttl_flights))
    print("--------------------------------------------------------------------")
    cat("Prop test for:", i)
    print(prop.test(c(delay_AM,delay_AK), c(ttl_AM,ttl_AK)))
}

```

Using a 5% level of significance,  we reject the null hypothesis of equal delay proportions for the following cities:  
  
* San Diego
* San Francisco
* Seattle  
  
We fail to reject the null hypothesis for the following cities:  
  
* Los Angeles
* Phoenix  
  
**In Aggregate**  
Finally, let's repeat the proportions test for the aggregated flight data:  
  
```{r, comment=NA}
flight_all_city2 <- flight %>%
                    group_by(airline) %>%
                    summarise(
                        delayed = sum(delayed),
                        ttl_flights = sum(ttl_flights)
                        )

delay_AM2 <- sum(select(filter(flight_all_city2, airline == 'AM WEST'),delayed))
delay_AK2 <- sum(select(filter(flight_all_city2, airline =='ALASKA'), delayed))
ttl_AM2 <- sum(select(filter(flight_all_city2, airline == 'AM WEST'),ttl_flights))
ttl_AK2 <- sum(select(filter(flight_all_city2, airline == 'ALASKA'),ttl_flights))

print(prop.test(c(delay_AM2,delay_AK2), c(ttl_AM2,ttl_AK2)))

```
  
At the 5% level of significance, we reject the null hypothesis that the difference in proportions is zero.  

